---
title: "2024 Data Call Outlier QA"
author: "Scarlet Sands-Bliss"
date: "2024-01-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
```

*Note: run this code following 2023-sdwis_04_samplingresults.Rmd in a project environment to have all data frames loaded.*

### **1. Visualizations**

#### Arsenic (mcl: 10ug/L)

*Flag over 300ug/L*

```{r}
# Histogram
subset_arsenic <- subset(wql_9, AnalyteCode == 1005)

# Boxplot
ggplot(subset_arsenic, aes(x = factor(1005), y = Concentration)) +
  geom_boxplot(outlier.shape = NA) +  # Suppress default outliers
  geom_jitter(position = position_jitter(width = 0.3), color = "red", size = 1) +  # Display outliers as red points
  geom_hline(yintercept = 10, linetype = "dashed", color = "darkblue", size = 0.8) +  
  labs(title = "Arsenic: Box Plot with Outliers",
       x = "AnalyteCode",
       y = "Concentration") +
  scale_x_discrete(breaks = factor(1005)) +  # Set x-axis to a single number
  theme_minimal()

ggsave("./figures-tables/plot_arsenic.png", plot = last_plot(), width = 8, height = 6, units = "in", dpi = 300)

# Examining summary stats for Arsenic
describe(wql_9$Concentration[wql_9$AnalyteCode == 1005])
table(wql_9$DetectionLimit[wql_9$AnalyteCode==1005]) # Detection limit is 2
# I would just assume the outlier > 800 should be removed

# Flag potential errors
arsenic_flag <- subset_arsenic %>% mutate(Flag = case_when(Concentration <= 100 ~ 0,
                                                    Concentration > 100 ~ 1))
```

#### Atrazine (mcl: 1ug/L)

*No flag data frame*

```{r}
# Histogram
subset_atrazine <- subset(wql_9, AnalyteCode == 2050)

# Boxplot
ggplot(subset_atrazine, aes(x = factor(2050), y = Concentration)) +
  geom_boxplot(outlier.shape = NA) +  # Suppress default outliers
  geom_jitter(position = position_jitter(width = 0.3), color = "red", size = 1) +  # Display outliers as red points
    geom_hline(yintercept = 1, linetype = "dashed", color = "darkblue", size = 0.8) +  
  labs(title = "Atrazine: Box Plot with Outliers",
       x = "AnalyteCode",
       y = "Concentration") +
  scale_x_discrete(breaks = factor(2050)) +  # Set x-axis to a single number
  theme_minimal()

ggsave("./figures-tables/plot_atrazine.png", plot = last_plot(), width = 8, height = 6, units = "in", dpi = 300)

# Examining summary stats for atrazine
describe(wql_9$Concentration[wql_9$AnalyteCode == 2050])
unique(subset_atrazine$Concentration) # Only value 0.25. Is this an issue?

unique(wql$result[wql$analyte_code == 2050])

# Atrazine inspection: all values in original data set were NA. Why?
atr_orig <- subset(wql, analyte_code == 2050)

# Flag potential errors - none here. All are zero. Confirming in original SDWIS file.
atrazine_test <- subset(wql_0, Analyte.Code == 2050)
table(atrazine_test$Result, useNA = "always")
table(atrazine_test$MCL[atrazine_test$Analyte.Code==2050]) # MCL is 1ug/L


```

#### DEHP (mcl: 4ug/L)

*No flag data frame*

```{r}
# Subset
subset_dehp <- subset(wql_9, AnalyteCode == 2039)

# Boxplot

ggplot(subset_dehp, aes(x = factor(2039), y = Concentration)) +
  geom_boxplot(outlier.shape = NA) +  # Suppress default outliers
  geom_jitter(position = position_jitter(width = 0.3), color = "red", size = 1) +  # Display outliers as red points
    geom_hline(yintercept = 4, linetype = "dashed", color = "darkblue", size = 0.8) +  
  labs(title = "DEHP: Box Plot with Outliers",
       x = "AnalyteCode",
       y = "Concentration") +
  scale_x_discrete(breaks = factor(2039)) +  # Set x-axis to a single number
  theme_minimal()

ggsave("./figures-tables/plot_DEHP.png", plot = last_plot(), width = 8, height = 6, units = "in", dpi = 300)

# Examining summary stats 
describe(wql_9$Concentration[wql_9$AnalyteCode == 2039])
table(wql_9$DetectionLimit[wql_9$AnalyteCode==2039]) # Detection limit is 3
print(min(subset_dehp$Concentration))
print(max(subset_dehp$Concentration))
table(subset_dehp$NonDetectFlag) # All but 4 were NDs


# Flag potential errors - none here. All are NDs except 4 that are all < 10. Confirming in original SDWIS file.
dehp_test <- subset(wql_0, Analyte.Code == 2039)
table(dehp_test$Result, useNA = "always")
table(dehp_test$MCL[dehp_test$Analyte.Code==2039]) # MCL is 4ug/L

```

#### Nitrate (1040, mcl: 10ug/L)

*Flags over 50ug/L*

```{r}
# Density Plot
subset_nitrate <- subset(wql_9, AnalyteCode==1040)

# Boxplot
ggplot(subset_nitrate, aes(x = factor(1040), y = Concentration)) +
  geom_boxplot(outlier.shape = NA) +  # Suppress default outliers
  geom_jitter(position = position_jitter(width = 0.3), color = "red", size = 1) +  # Display outliers as red points
    geom_hline(yintercept = 10, linetype = "dashed", color = "darkblue", size = 0.8) +  
  labs(title = "Nitrate: Box Plot with Outliers",
       x = "AnalyteCode",
       y = "Concentration") +
  scale_x_discrete(breaks = factor(1040)) +  # Set x-axis to a single number
  theme_minimal()

ggsave("./figures-tables/plot_nitrate.png", plot = last_plot(), width = 8, height = 6, units = "in", dpi = 300)

# Examining summary stats
describe(wql_9$Concentration[wql_9$AnalyteCode %in% nitrate_as_n_code])
print(min(subset_nitrate$Concentration))
print(max(subset_nitrate$Concentration))

# Flag potential errors
nitrate_flag <- subset_nitrate %>% mutate(Flag = case_when(Concentration <= 50 ~ 0,
                                                    Concentration > 50 ~ 1))

# Several high balues in the 50-80 range
```

#### PCE (mcl: 5 ug/L)

*Flag for \> 100, but many high values ranging from 5-200*

```{r}
# Density Plot
subset_pce <- subset(wql_9, AnalyteCode == 2987)

# Boxplot

ggplot(subset_pce, aes(x = factor(2987), y = Concentration)) +
  geom_boxplot(outlier.shape = NA) +  # Suppress default outliers
  geom_jitter(position = position_jitter(width = 0.3), color = "red", size = 1) +  # Display outliers as red points
    geom_hline(yintercept = 5, linetype = "dashed", color = "darkblue", size = 0.8) +  
  labs(title = "PCE: Box Plot with Outliers",
       x = "AnalyteCode",
       y = "Concentration") +
  scale_x_discrete(breaks = factor(2987)) +  # Set x-axis to a single number
  theme_minimal()

ggsave("./figures-tables/plot_PCE.png", plot = last_plot(), width = 8, height = 6, units = "in", dpi = 300)

# Examining summary stats
describe(wql_9$Concentration[wql_9$AnalyteCode == 2987])
table(wql_9$DetectionLimit[wql_9$AnalyteCode==2987]) # Detection limit is 0.5
print(min(subset_pce$Concentration))
print(max(subset_pce$Concentration))
table(subset_pce$NonDetectFlag) # Many non-detects.

# Examining original SDWIS file
pce_test <- subset(wql_0, Analyte.Code == 2987)
table(pce_test$Result, useNA = "always") # Majority ND
table(pce_test$MCL[pce_test$Analyte.Code==2987]) # MCL is 5 ug/L

# Flag potential errors
pce_flag <- subset_pce %>% mutate(Flag = case_when(Concentration <= 100 ~ 0,
                                                    Concentration > 100 ~ 1))

# Many high values in 25-200 range.



```

#### Radium (Combined 226 & 228, mcl: 5ug/L)

*No flag data frame*

```{r}
# Density Plot
subset_rad <- subset(wql_9, AnalyteCode == 4010)

# Boxplot
ggplot(subset_rad, aes(x = factor(4010), y = Concentration)) +
  geom_boxplot(outlier.shape = NA) +  # Suppress default outliers
  geom_jitter(position = position_jitter(width = 0.3), color = "red", size = 1) +  # Display outliers as red points
    geom_hline(yintercept = 5, linetype = "dashed", color = "darkblue", size = 0.8) +  
  labs(title = "Radium: Box Plot with Outliers",
       x = "Analyte Code",
       y = "Concentration") +
  scale_x_discrete(breaks = factor(4010)) +  # Set x-axis to a single number
  theme_minimal()

ggsave("./figures-tables/plot_radium.png", plot = last_plot(), width = 8, height = 6, units = "in", dpi = 300)

# Examining summary stats
describe(wql_9$Concentration[wql_9$AnalyteCode == 4010])
print(min(subset_rad$Concentration))
print(max(subset_rad$Concentration))
table(subset_rad$NonDetectFlag) # Al but 4 were non-detects. 

# Examining original SDWIS file
rad_test <- subset(wql_0, Analyte.Code == 4010)
table(rad_test$Result, useNA = "always") # Majority ND
table(rad_test$MCL[rad_test$Analyte.Code==4010]) # MCL is 5 ug/L

```

#### TCE (mcl: 5ug/L)

*Flag for \> 100ug/L*

```{r}
# Density Plot
subset_tce <- subset(wql_9, AnalyteCode == 2984)

# Boxplot
ggplot(subset_tce, aes(x = factor(2984), y = Concentration)) +
  geom_boxplot(outlier.shape = NA) +  # Suppress default outliers
  geom_jitter(position = position_jitter(width = 0.3), color = "red", size = 1) +  # Display outliers as red points
    geom_hline(yintercept = 5, linetype = "dashed", color = "darkblue", size = 0.8) +  
  labs(title = "TCE: Box Plot with Outliers",
       x = "Analyte Code",
       y = "Concentration") +
  scale_x_discrete(breaks = factor(2984)) +  # Set x-axis to a single number
  theme_minimal()

ggsave("./figures-tables/plot_TCE.png", plot = last_plot(), width = 8, height = 6, units = "in", dpi = 300)

# Examining summary stats
describe(wql_9$Concentration[wql_9$AnalyteCode == 2984])
print(min(subset_tce$Concentration))
print(max(subset_tce$Concentration))
table(subset_tce$NonDetectFlag) # More detect than ND

# Examining original SDWIS file
tce_test <- subset(wql_0, Analyte.Code == 2984)
table(tce_test$Result, useNA = "always") 
table(tce_test$MCL[tce_test$Analyte.Code==2984]) # MCL is 5 ug/L

# Flag potential errors
tce_flag <- subset_tce %>% mutate(Flag = case_when(Concentration <= 100 ~ 0,
                                                    Concentration > 100 ~ 1))
# Range from 0.25 to 390, decent number of scattered values over 100

```

#### Uranium (mcl: 30ug/L)

*Flag for \> 200ug/L*

```{r}
# Histogram
subset_uranium <- subset(wql_9, AnalyteCode == 4006)

# Box plot
ggplot(subset_uranium, aes(x = factor(4006), y = Concentration)) +
  geom_boxplot(outlier.shape = NA) +  # Suppress default outliers
  geom_jitter(position = position_jitter(width = 0.3), color = "red", size = 1) +  # Display outliers as red points
    geom_hline(yintercept = 30, linetype = "dashed", color = "darkblue", size = 0.8) +  
  labs(title = "Uranium: Box Plot with Outliers",
       x = "AnalyteCode",
       y = "Concentration") +
  scale_x_discrete(breaks = factor(4006)) +  # Set x-axis to a single number
  theme_minimal()

ggsave("./figures-tables/plot_uranium.png", plot = last_plot(), width = 8, height = 6, units = "in", dpi = 300)

# Examining summary stats for Uranium
describe(wql_9$Concentration[wql_9$AnalyteCode == 4006])

# Examining original SDWIS file
uranium_test <- subset(wql_3, analyte_code == 4006)
table(uranium_test$mcl[uranium_test$analyte_code==4006]) # MCL is 5 ug/L

# Flag potential errors
uranium_flag <- subset_uranium %>% mutate(Flag = case_when(Concentration <= 100 ~ 0,
                                                    Concentration > 100 ~ 1))

# Many values over 200, not sure about. The value near 1000 is likely error.



```

### 2. Merge to potential outliers List

```{r}
# Create a list of data frames
list_of_dataframes <- list(arsenic_flag, nitrate_flag, uranium_flag)

# Bind the rows of all data frames together
outliers <- bind_rows(list_of_dataframes)

# Filter rows where Flag is 1
outliers <- outliers %>%
  filter(Flag == 1) %>%
  select(RowIdentifier,
         PWSIDNumber,
         Year,
         AnalyteCode,
         ConcentrationUnits,
         Concentration,
         DateSampled,
         SamplePointID,
         DetectionLimit,
         NonDetectFlag)

table(outliers$SamplePointID) 
length(unique(outliers$SamplePointID)) # All values just from 22 sampling points


# Save as CSV
write.csv(outliers, sprintf("data/clean-data/extreme-values_%i.csv", year_for_datacall)) 
```
